Supplemental Material: Multidimensional analysis and detection of informative features in human brain white matter

### 2021-06-07 reports/manuscript/paper_changes.tex

@@

Using data from a previous study of patients with amyotrophic lateral sclerosis (ALS)~\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on \emph{a priori}
selection of features only within the CST bundle-of-interest.
SGL delivers improved predictive performance, with a cross-validated accuracy
of {\alsAccuracy}\% and an area under the receiver operating characteristic
curve (ROC AUC) of {\alsRocAuc}, without the need for \emph{a priori} feature
engineering.
\added{We also predicted ALS diagnosis using the PCR-SGL, wherein each group of features is independently transformed to its PC basis and an SGL model is fit to the transformed features. This model achieves \protect{\alsAccuracyGpca}\% accuracy and an ROC AUC of \protect{\alsRocAucGpca}}
(\cref{fig:class-results:probs}). In addition to this classification
performance, both SGL \added{and PCR-SGL} also \replaced{identify}{identifies}
the white matter tracts most important for ALS classification.
\added{\Cref{fig:class-results:coefs3d} shows that PCR-SGL identified as potential disease biomarkers the diffusion measures in the CST from the cerebral peduncle to the corona radiata, agreeing with the previous study from which these data were extracted \mbox{\cite{sarica2017corticospinal}}.}
The relative importance of white matter
features is captured in the $\beta$ coefficients from \cref{eq:sgl}.
\Cref{fig:class-results:tract-profiles} depicts these coefficients along the
right CST, plotted over the FA values for the control and ALS subject groups \added{(see supplemental Figs S1 and S2 for tract profiles of all 18 tracts in these groups)}.
We find that SGL \replaced{and PCR-SGL select}{selects} FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings \cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in
the previous study from which we obtained the data
\cite{sarica2017corticospinal}.

@@

@@

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 1651, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable $y$. SGL was fit to the \replaced{tract profile}{tractometry-extracted}
features FA and mean diffusivity (MD) in 18 major brain tracts, with diffusion metrics extracted
from diffusion tensor imaging (DTI) for the WH dataset and diffusion kurtosis
imaging (DKI) \cite{jensen2005diffusion} for the HBN and Cam-CAN datasets \added{see supplemental Figs S3 to S8 for the full tract profile information in all tracts/datasets}.

@@

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
\added{As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.}
In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
\added{We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the supplementary material. For age regression, substantial differences between the datasets preclude generalization across all three:}
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item \added{different acquisition parameters, which can be challenging to harmonize \mbox{\cite{pinto2020harmonization}}}
    \item \added{different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets}
    \item \added{different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets}
    \item \added{different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).}
\end{enumerate*}
\added{In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \mbox{\cite{lermausabiaga2019replication}}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.}

@@

\item Diffusion MRI data from the Healthy Brain Network pediatric mental
health study \cite{alexander2017open}, containing dMRI data from 1651 subjects
with ages 5-21. These data were measured in 3T Siemens MRI scanners at two
sites in the New York
area\added{: Rutgers University Brain Imaging Center and the CitiGroup Cornell Brain Imaging Center}.
Informed consent was obtained from each participant aged 18 or older. For
participants younger than 18, written consent was obtained from their legal
guardians and written assent was obtained from the participant. Voxel
resolution was \num{1.8x1.8x1.8}$\text{mm}^3$ with 64 non-colinear directions
measured for each of $b=1000 \text{s} / \text{mm}^2$ and $b=2000 \text{s} /
\text{mm}^2$. Preprocessing was performed using \emph{QSIPrep} 0.12.1, which
is based on \emph{Nipype} 1.5.1 \cite[RRID:SCR\_002502]{nipype1,nipype2}.

@@

Data from the ALS and WH studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox
(\texttt{mAFQ}\added{, version 1.1 for WH and version 1.2 for ALS})
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. \replaced{Eighteen major}{Major}
tracts\added{, which are enumerated in the supplemental material,} were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a bundle profile of tissue properties in each bundle was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
bundle. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
\replaced{mahalanobis}{mahalnobis} distance of the node from the average of
that node across streamlines. This means that streamlines that are more
representative of the tract contribute more to the bundle profile, relative
to streamlines that are on the edge of the tract.

@@

These processes create bundle profiles, in which diffusion measures
are quantified and averaged along eighteen major fiber tracts,
which are enumerated in the supplemental material.
\added{See Fig S3 of \mbox{\cite{kruper2021evaluating}} for a depiction of these white matter bundles.}
Here,
we use only the mean diffusivity (MD) and the fractional anisotropy
(FA) of the diffusion tensor, but additional dMRI-based maps or maps
based on other quantitative MRI measurements can also be used.
\added{The resulting feature space was the same for all four datasets, with the FA and MD metrics at each of 100 nodes in eighteen bundles comprising 3600 features per subject.}
These bundle profiles, along with the phenotypical data we wish to explain
or predict, form the input to the SGL algorithm. In a domain-agnostic
machine learning context, the phenotypical data comprise the target
variables while the bundle profiles form the feature or predictor
variables (\replaced{see}{See} \cref{fig:methods:group-structure}).

@@

\added{For the multisite HBN study, we use the ComBat harmonization method to robustly adjust for site effects in the tract profiles. Initially designed to correct for site effects in gene expression studies \mbox{\cite{johnson2007adjusting}}, ComBat employs a parametric empirical Bayes approach to adjust for batch effects and has since been applied to multi-site cortical thickness measurements \mbox{\cite{fortin2018harmonization}}, multi-site DTI studies \mbox{\cite{fortin2017harmonization}}, and functional MRI data in the Adolescent Brain Cognitive Development Study (ABCD) \mbox{\cite{nielson2018detecting}}. We rely on the \texttt{neurocombat\_sklearn} library \mbox{\cite{neurocombat_sklearn}}, to apply ComBat in our scikit-learn analysis pipeline and present bundle profile site differences in the supplemental material.}

@@

@@

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Appendix.}
\label{S1_Appendix}
{\bf Bundle and regression coefficient profiles.}
In this supporting section, we present bundle profiles and $\hat{\beta}$ coefficients
for each anatomical bundle and in each dataset.

@@

\newpage

\section*{Figure legends}

\begin{figure}[b!]
    % \vspace{0.25cm}
    % \rlap{
    %     \hspace{-0.25em}
    %     \includegraphics[width=0.98\columnwidth]{methods-brains.pdf}
    % }
    % \vspace{-8.4cm}
    % \includegraphics[width=\columnwidth]{method-quad.pdf}
    {\phantomsubcaption\label{fig:methods:tractogram}}
    {\phantomsubcaption\label{fig:methods:cst}}
    {\phantomsubcaption\label{fig:methods:tract-profile:cst}}
    {\phantomsubcaption\label{fig:methods:tract-profile:fa}}
    {\phantomsubcaption\label{fig:methods:group-structure}}
    \phantomcaption{\label{fig:methods}}
\end{figure}

\begin{figure}[b!]
    % \vspace{3.0cm}
    % \rlap{
    %     \includegraphics[width=0.33\textwidth]{als_gpca_coefs.pdf}
    % }
    % \vspace{-7.0cm}
    % \includegraphics[width=\textwidth]{als_gpca_results.pdf}
    {\phantomsubcaption\label{fig:class-results:probs}}
    {\phantomsubcaption\label{fig:class-results:coefs3d}}
    {\phantomsubcaption\label{fig:class-results:tract-profiles}}
    {\phantomsubcaption\label{fig:class-results:false-positives}}
    {\phantomsubcaption\label{fig:class-results:false-negatives}}
    \phantomcaption{%
    \label{fig:class-results}
    }
\end{figure}

\begin{figure}[b!]
    % \vspace{3.0cm}
    % \rlap{
    %     \hspace{0.775cm}
    %     \includegraphics[width=0.26\textwidth]{wh_coefs.pdf}
    %     \hspace{0.5em}
    %     \includegraphics[width=0.26\textwidth]{hbn_coefs.pdf}
    %     \hspace{0.5em}
    %     \includegraphics[width=0.26\textwidth]{camcan_coefs.pdf}
    % }
    % \vspace{-6.1cm}
    % \includegraphics[width=\textwidth]{regression_scatter.pdf}
    {\phantomsubcaption\label{fig:age-results:wh-scatter}}
    {\phantomsubcaption\label{fig:age-results:hbn-scatter}}
    {\phantomsubcaption\label{fig:age-results:cc-scatter}}
    {\phantomsubcaption\label{fig:age-results:wh-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:hbn-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:cc-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:wh-profile}}
    {\phantomsubcaption\label{fig:age-results:hbn-profile}}
    {\phantomsubcaption\label{fig:age-results:cc-profile}}
    \phantomcaption{%
    \label{fig:age-results}
    }
\end{figure}

\begin{figure}[b!]
    % \includegraphics[width=\textwidth]{model_performance.pdf}
    \phantomcaption{%
    \label{fig:model-performance}
    }
\end{figure}

\begin{figure}[h!]
    % \includegraphics[width=0.75\textwidth]{nested-cross-validation.pdf}
    \phantomcaption{
    \label{fig:nested-cross-val}
    }
\end{figure}

>>>>

\begin{figure}[b!]
    % \vspace{0.25cm}
    % \rlap{
    %     \hspace{-0.25em}
    %     \includegraphics[width=0.98\columnwidth]{methods-brains.pdf}
    % }
    % \vspace{-8.4cm}
    % \includegraphics[width=\columnwidth]{method-quad.pdf}
    {\phantomsubcaption\label{fig:methods:tractogram}}
    {\phantomsubcaption\label{fig:methods:cst}}
    {\phantomsubcaption\label{fig:methods:tract-profile:cst}}
    {\phantomsubcaption\label{fig:methods:tract-profile:fa}}
    {\phantomsubcaption\label{fig:methods:group-structure}}
    \caption{\label{fig:methods}
    {\bf Figure 1: Tractometry data flow}
    \textbf{(a)} Whole brain tractography generates streamlines approximating
        the trajectories of white matter connections.
        \textbf{(b)} Tractometry classifies these streamlines into anatomical bundles.
        In this case, we show the left corticospinal tract (CSTL)
        \added{and the left arcuate fasciculus (ARCL)}
        over a mid-saggital anatomical slice.
        \deleted{\textbf{(c)}} \replaced{Tract profiling}{Tractometry} further extracts bundle profiles,
        quantifications of various diffusion metrics along the length of the
        fiber bundle. Here, we show one subject's fractional anisotropy (FA)
        profile for \added{\textbf{(c)}} the CSTL \added{and \textbf{(d)} the ARCL}.
        \textbf{(d)} the phenotypical target data and \replaced{tract profile}{tractometric}
        features can be organized into a linear model, $\hat{y} = \mathbf{X}
        \hat{\beta}$. The feature matrix $\mathbf{X}$ is color-coded
        to reveal a natural group structure: the left (orange) group
        contains $k$ features from the CSTL, the middle (green) group
        contains $k$ features from the left cingulum cingulate (CGCL),
        and the right (blue) group
        contains $k$ features from the \replaced{ARCL}{left arcuate (ARCL)}.
        The coefficients in $\hat{\beta}$ follow the same natural grouping. \added{Panels (a) and (b) are adapted from \url{https://figshare.com/articles/figure/example_tractography-segmentation/14485350}, and reproduced under the CC-BY license (\url{https://creativecommons.org/licenses/by/4.0/})}
    }
\end{figure}

@@

Using data from a previous study of patients with amyotrophic lateral sclerosis (ALS)~\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on \emph{a priori}
selection of features only within the CST bundle-of-interest.
SGL delivers improved predictive performance, with a cross-validated accuracy
of {\alsAccuracy}\% and an area under the receiver operating characteristic
curve (ROC AUC) of {\alsRocAuc}, without the need for \emph{a priori} feature
engineering.
\added{We also predicted ALS diagnosis using the PCR-SGL, wherein each group of features is independently transformed to its PC basis and an SGL model is fit to the transformed features. This model achieves \protect{\alsAccuracyGpca}\% accuracy and an ROC AUC of \protect{\alsRocAucGpca}}
(\cref{fig:class-results:probs}). In addition to this classification
performance, both SGL \added{and PCR-SGL} also \replaced{identify}{identifies}
the white matter tracts most important for ALS classification.
\added{\Cref{fig:class-results:coefs3d} shows that PCR-SGL identified as potential disease biomarkers the diffusion measures in the CST from the cerebral peduncle to the corona radiata, agreeing with the previous study from which these data were extracted \mbox{\cite{sarica2017corticospinal}}.}
The relative importance of white matter
features is captured in the $\beta$ coefficients from \cref{eq:sgl}.
\Cref{fig:class-results:tract-profiles} depicts these coefficients along the
right CST, plotted over the FA values for the control and ALS subject groups \added{(see Supporting Figures S1 and S2 for tract profiles of all 18 tracts in these groups)}.
We find that SGL \replaced{and PCR-SGL select}{selects} FA metrics in the corticospinal tract and
particularly in the right corticospinal tract as most important to ALS
classification, confirming previous findings \cite{van2011upper,
toosy2003diffusion, sarica2014tractography, sage2007quantitative,
sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion,
cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and
identifying the portions of the brain that were selected \emph{a priori} in
the previous study from which we obtained the data
\cite{sarica2017corticospinal}.

@@

\begin{figure}[b!]
    % \vspace{3.0cm}
    % \rlap{
    %     \includegraphics[width=0.33\textwidth]{als_gpca_coefs.pdf}
    % }
    % \vspace{-7.0cm}
    % \includegraphics[width=\textwidth]{als_gpca_results.pdf}
    {\phantomsubcaption\label{fig:class-results:probs}}
    {\phantomsubcaption\label{fig:class-results:coefs3d}}
    {\phantomsubcaption\label{fig:class-results:tract-profiles}}
    {\phantomsubcaption\label{fig:class-results:false-positives}}
    {\phantomsubcaption\label{fig:class-results:false-negatives}}
    \caption{
    {\bf Figure 2: \added{PCR-}SGL accurately and interpretably predicts ALS diagnosis.}
    {\bf (a)} Classification probabilities for ALS diagnosis, with
    controls on the left, patients on the right, predicted controls in
    blue, and predicted patients in orange. That is, orange dots on the left
    represent false positives, while blue dots on the right represent
    false negatives. We achieve {\protect\alsAccuracy}\% accuracy with an
    ROC AUC of {\protect\alsRocAuc}.
    {\bf (b)} \added{PCR-}SGL coefficients are presented on the core fibers of major
    fiber bundles. They exhibit high group sparsity and are concentrated
    in the FA of the \replaced{corticospinal tract (CST)}{CST}.
    The brain is oriented with the right hemisphere
    in the foreground and anterior to the right of the page. The CSTL, CSTR,
    \replaced{callosum forceps anterior (CFA), left arcuate (ARCL), and right arcuate (ARCR)}{CFA}
    bundles are indicated for orientation.
    {\bf (c)} \added{PCR-}SGL identifies three portions of the CST as important,
    where $\hat{\beta}$ (dashed line, right axis) has large values. These
    are centered around nodes 30, 65, and 90, corresponding to locations
    of substantial differences in FA between the ALS and control groups
    (shaded areas indicates standard error of the mean).
    {\bf (d)} Bundle profiles for false positive classifications. Line
    colors correspond to the marker edge color in the top left plot.
    These individuals have reduced FA in the CST portions which SGL
    identified as important. Their misclassification is coherent with the
    feature importance and the group differences in FA.
    {\bf (e)} Individual bundle profiles for false negative
    classifications. These individuals have bundle profiles which
    oscillate between the group means.
    \label{fig:class-results}
    }
\end{figure}

@@

The WH \cite{yeatman2014lifespan}, HBN \cite{alexander2017open}, and Cam-CAN
\cite{shafto2014cambridge,taylor2017cambridge} datasets used here contain
data from 76, 1651, and 640 subjects, respectively, ranging from 6-50, 5-21,
and 18-88 years of age, respectively. In each case, biological age was used
as the predicted variable $y$. SGL was fit to the \replaced{tract profile}{tractometry-extracted}
features FA and mean diffusivity (MD) in 18 major brain tracts, with diffusion metrics extracted
from diffusion tensor imaging (DTI) for the WH dataset and diffusion kurtosis
imaging (DKI) \cite{jensen2005diffusion} for the HBN and Cam-CAN datasets \added{see Supporting Figures S3 to S8 for the full tract profile information in all tracts/datasets}.

@@

\begin{figure}[b!]
    % \vspace{3.0cm}
    % \rlap{
    %     \hspace{0.775cm}
    %     \includegraphics[width=0.26\textwidth]{wh_coefs.pdf}
    %     \hspace{0.5em}
    %     \includegraphics[width=0.26\textwidth]{hbn_coefs.pdf}
    %     \hspace{0.5em}
    %     \includegraphics[width=0.26\textwidth]{camcan_coefs.pdf}
    % }
    % \vspace{-6.1cm}
    % \includegraphics[width=\textwidth]{regression_scatter.pdf}
    {\phantomsubcaption\label{fig:age-results:wh-scatter}}
    {\phantomsubcaption\label{fig:age-results:hbn-scatter}}
    {\phantomsubcaption\label{fig:age-results:cc-scatter}}
    {\phantomsubcaption\label{fig:age-results:wh-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:hbn-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:cc-coefs3d}}
    {\phantomsubcaption\label{fig:age-results:wh-profile}}
    {\phantomsubcaption\label{fig:age-results:hbn-profile}}
    {\phantomsubcaption\label{fig:age-results:cc-profile}}
    \phantomcaption{%
    {\bf Figure 3: Predicting age with tractometry and SGL.}
    {\bf (top)} The predicted age vs. true age of each individual from the test
splits (i.e., when each subject's data was held out in fitting the model) for
the {\bf (a)} WH, {\bf (b)} HBN, and {\bf (c)} Cam-CAN datasets; an accurate
prediction falls close to the $y=x$ line (dashed). The mean absolute error (MAE)
and coefficient of determination $R^2$ are presented in the lower right of each
scatter plot. {\bf (middle)} Feature importance for predicting age from
\replaced{tract profile}{tractometry} in the {\bf (d)} WH, {\bf (e)} HBN, and
{\bf (f)} Cam-CAN datasets. The orientation of the brain is that same as in
\cref{fig:class-results:coefs3d}, however because the coefficients exhibit high
global sparsity (as opposed to group sparsity), we plot the mean of the absolute
value of $\hat{\beta}$ for each bundle on the core fiber. The global
distrubution of the $\hat{\beta}$ coefficients reflects the fact that aging is
not confined to a single white matter bundle. {\bf (bottom)} Age quintile bundle
profiles for the {\bf (g)} WH, {\bf (h)} HBN, and {\bf (i)} Cam-CAN datasets.
    \label{fig:age-results}
    }
\end{figure}

\begin{figure}[b!]
    % \includegraphics[width=\textwidth]{model_performance.pdf}
    \caption{%
    {\bf \added{Figure 4: Model performance across all datasets.}}
\added{Each panel shows model performance measured on the test set for each}
\added{cross-validation split, with each black dot representing a split, box}
\added{plots representing the quartiles, and white diamonds representing the mean}
\added{performance. The $y$-scale varies in each subplot.}
\added{\textbf{(a)} Accuracy of test set predictions for the}
\added{ALS dataset. Because group differences}
\added{in ALS diagnosis are mostly confined to a single bundle, the group}
\added{structure-preserving methods, SGL and PCR-SGL, outperform the other}
\added{models.}
\added{The remaining frames show coefficient of determination, \protect$R^2$ in test sets}
\added{for the \textbf{(b)} WH, \textbf{(c)} HBN, and \textbf{(c)} Cam-CAN}
\added{datasets. Because aging affects the white matter globally, group structure-blind}
\added{methods like elastic net and PCR Lasso perform well. Nonetheless, the SGL}
\added{models show competitve predictive performance, adapting to a problem where}
\added{group structure is not as informative. PCR-SGL performs poorly in this regime}
\added{because its initial group-wise PC projection destroys between bundle covariance.}
\added{The bundle-mean lasso performs poorly, demonstrating the value of along-tract profiling.}
    \label{fig:model-performance}
    }
\end{figure}

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
\added{As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.}
In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
\added{We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the Supporting Information. For age regression, substantial differences between the datasets preclude generalization across all three:}
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item \added{different acquisition parameters, which can be challenging to harmonize \mbox{\cite{pinto2020harmonization}}}
    \item \added{different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets}
    \item \added{different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets}
    \item \added{different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).}
\end{enumerate*}
\added{In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \mbox{\cite{lermausabiaga2019replication}}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.}

@@

\item Diffusion MRI data from the Healthy Brain Network pediatric mental
health study \cite{alexander2017open}, containing dMRI data from 1651 subjects
with ages 5-21. These data were measured in 3T Siemens MRI scanners at two
sites in the New York
area\added{: Rutgers University Brain Imaging Center and the CitiGroup Cornell Brain Imaging Center. Age distributions of participants in the different sites are in Supporting Figure S9}.
Informed consent was obtained from each participant aged 18 or older. For
participants younger than 18, written consent was obtained from their legal
guardians and written assent was obtained from the participant. Voxel
resolution was \num{1.8x1.8x1.8}$\text{mm}^3$ with 64 non-colinear directions
measured for each of $b=1000 \text{s} / \text{mm}^2$ and $b=2000 \text{s} /
\text{mm}^2$. Preprocessing was performed using \emph{QSIPrep} 0.12.1, which
is based on \emph{Nipype} 1.5.1 \cite[RRID:SCR\_002502]{nipype1,nipype2}.

@@

Data from the ALS and WH studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox
(\texttt{mAFQ}\added{, version 1.1 for WH and version 1.2 for ALS})
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. \replaced{Eighteen major}{Major}
tracts\added{, which are enumerated in Figure S1,} were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a bundle profile of tissue properties in each bundle was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
bundle. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
\replaced{mahalanobis}{mahalnobis} distance of the node from the average of
that node across streamlines. This means that streamlines that are more
representative of the tract contribute more to the bundle profile, relative
to streamlines that are on the edge of the tract.

@@

These processes create bundle profiles, in which diffusion measures
are quantified and averaged along eighteen major fiber tracts,
which are enumerated in the Supporting Information.
\added{See Fig S3 of \mbox{\cite{kruper2021evaluating}} for a depiction of these white matter bundles.}
Here,
we use only the mean diffusivity (MD) and the fractional anisotropy
(FA) of the diffusion tensor, but additional dMRI-based maps or maps
based on other quantitative MRI measurements can also be used.
\added{The resulting feature space was the same for all four datasets, with the FA and MD metrics at each of 100 nodes in eighteen bundles comprising 3600 features per subject.}
These bundle profiles, along with the phenotypical data we wish to explain
or predict, form the input to the SGL algorithm. In a domain-agnostic
machine learning context, the phenotypical data comprise the target
variables while the bundle profiles form the feature or predictor
variables (\replaced{see}{See} \cref{fig:methods:group-structure}).

@@

\added{For the multisite HBN study, we use the ComBat harmonization method to robustly adjust for site effects in the tract profiles. Initially designed to correct for site effects in gene expression studies \mbox{\cite{johnson2007adjusting}}, ComBat employs a parametric empirical Bayes approach to adjust for batch effects and has since been applied to multi-site cortical thickness measurements \mbox{\cite{fortin2018harmonization}}, multi-site DTI studies \mbox{\cite{fortin2017harmonization}}, and functional MRI data in the Adolescent Brain Cognitive Development Study (ABCD) \mbox{\cite{nielson2018detecting}}. We rely on the \texttt{neurocombat\_sklearn} library \mbox{\cite{neurocombat_sklearn}}, to apply ComBat in our scikit-learn analysis pipeline and present bundle profile site differences and ComBat correction in Supporting Figures S10-S13.}

@@

\begin{figure}[h!]
    % \includegraphics[width=0.75\textwidth]{nested-cross-validation.pdf}
    \phantomcaption{
    {\bf Figure 5: Nested cross-validation.}
We evaluate model quality using a nested $k$-fold cross validation
scheme. At level-0, the input data is decomposed into $k_0$ shuffled
groups and optimal hyperparameters are found for the level-0 training
set. To avoid overfitting, the optimal hyperparameters are themselves
evaluated using a cross-validation scheme taking place at level-1 of
the decomposition, where each level-0 training set is further
decomposed into $k_1 = 3$ shuffled groups. In the classification
case, the training and test splits are stratified by diagnosis. For
the ALS and WH data, $k_0 = 10$, while for the HBN and Cam-CAN data,
$k_0 = 5$.
    \label{fig:nested-cross-val}
    }
\end{figure}

@@

@@

\paragraph{Figure S1: Bundle profiles and regression coefficients for the ALS dataset FA.}
Diffusion metrics are plotted
along the length of eighteen bundles:
right corticospinal (CSTR),
left corticospinal (CSTL),
right uncinate (UNCR),
left uncinate (UNCL),
left inferior fronto-occipital fasciculus (IFOL),
right inferior fronto-occipital fasciculus (IFOR),
right arcuate (ARCR),
left arcuate (ARCL),
right thalamic radiation (ATRR),
left thalamic radiation (ATRL),
right cingulum cingulate (CGCR),
left cingulum cingulate (CGCL),
callosum forceps posterior (CFP),
callosum forceps anterior (CFA),
right inferior longitudinal fasciculus (ILFR),
left inferior longitudinal fasciculus (ILFL),
right superior longitudinal fasciculus (SLFR),
and left superior longitudinal fasciculus (SLFL). We use these acronyms throughout the Supporting Information.
FA is plotted on the left $y$-axis while the $\hat{\beta}$ coefficients are displayed on the twin axis on the right-hand-side. SGL selected
the right corticospinal tract (CSTR) as important and regularized coefficients
in the CSTL. Yet, there are also group FA differences in the
CSTL. This highlighted a potential drawback of the SGL method, discussed in the
main text in the context of age regression. Namely, SGL is not guaranteed to
identify \emph{all} important features. In this case, if the diagnostic signal
in the CSTL is redundant to that in the CSTR, SGL will regularize the CSTL
features, thereby reducing its sparsity penalty without any corresponding
increase in loss. This parsimony cuts both ways; it is a feature of the method
when one seeks an efficient predictive model, but is a disadvantage of the
method when one wants an exhaustive explanation of feature importance. We use the phrase ``parsimony pitfall'' to refer to the case when SGL regularizes
away redundant but obviously important features.

\paragraph*{Figure S2: Bundle profiles and regression coefficients for the ALS dataset MD.} The scale of the $\hat{\beta}$-axis is identical to that used in Figure S1, to facilitate the comparison of the relative importance of each metric.

\paragraph*{Figure S3: Bundle profiles and regression
coefficients for the WH dataset FA.} In contrast to the ALS classification case,
the $\hat{\beta}$ coefficients are distributed widely through the brain, supporting the interpretation that aging is a large and continuous whole-brain process.

\paragraph*{Figure S4: Bundle profiles and regression coefficients for the WH
dataset MD.} This and Figure S3 demonstrate that SGL behaves much more like the
lasso than the group lasso, as discussed in the main text. The parsimony pitfall
is most evident in the IFOL and IFOR bundles.

\paragraph*{Figure S5: FA bundle profiles and $\hat{\beta}$ coefficients for age
regression in the HBN dataset.} Like the WH dataset, the $\hat{\beta}$
coefficients are distributed widely through the brain and SGL behaves more like
the lasso than the group lasso.

\paragraph*{Figure S6: MD bundle profiles and $\hat{\beta}$ coefficients for age
regression in the HBN dataset.} In contrast to the WH results, the bundle
profiles show different behaviors. For example the SLFL and SLFR bundle profiles
in have different concavity. This is unsurprising, however, given the
differences between these datasets (see also Discussion). The parsimony pitfall
is most evident in the UNCL, UNCR, ARCL, SLFL, and SLFR bundles

\paragraph*{Figure S7: FA bundle profiles and $\hat{\beta}$ coefficients for age
regression in the Cam-CAN dataset.} The $\hat{\beta}$ coefficients are
distributed widely through the brain and SGL behaves more like the lasso than
the group lasso. As before, one must be cautious about comparing bundle profiles
and $\hat{\beta}$ coefficients between models. While the HBN and Cam-CAN
datasets share the same diffusion model and refrain from clipping streamlines,
the age distributions for the two are roughly disjoint, with the WH age
distribution straddling the two.

\paragraph*{Figure S8: Mean diffusivity (MD) bundle profiles and $\hat{\beta}$
coefficients for age regression in the Cam-CAN dataset.}

\paragraph*{Figure S9: Age distributions are similar between the different
HBN sites.} Rutgers Rutgers University Brain Imaging Center (RU) and the CitiGroup Cornell Brain Imaging Center (CBIC).

\paragraph*{Figure S10: FA bundle profiles exhibit strong site differences in the HBN dataset.}

\paragraph*{Figure S11: MD bundle profiles exhibit strong site differences in the HBN dataset.}

\paragraph*{Figure S12: Site differences in FA are eliminated by ComBat harmonization.}

\paragraph*{Figure S13: Site differences in MD are eliminated by ComBat harmonization.}
